A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-21 tex/codatadp.tex
Split report/paper versions.  Minor.

Given a $\mu$-term $t$ of the form $\MU x.\; \const{C}( t_1, \ldots, t_n )$ with subterm $u$,
we say $u$ is a \emph{bisimilar subterm} of $t$ if it is of the form
$\MU y.\; \const{C}( u_1, \ldots, u_n )$ and $t\interp{ t_k } \aequiv t\interp{ u_k }$ for $k = 1,\ldots, n$.
The term $t$ is \emph{normal} if it does not contain a strict, bisimilar subterm
and all of its strict subterms are also normal.
For example, the $\mu$-term $t = \MU x.\; \const{C}( \MU y.\; \const{C}( y ) )$ is abnormal,
since $\MU y.\; \const{C}( y )$ is a bisimilar subterm of $t$.
In particular, note that their arguments have the same interpretation in $t$,
that is,
$t\interp{\MU y.\; \const{C}( y )}^\emptyset =$
$\MU y.\; \const{C}( t\interp{y}^{\{y\}} ) =$
$\MU y.\; \const{C}( y )$
is $\alpha$-equivalent to
$t\interp{y}^\emptyset =$
$\MU y.\; \const{C}( t\interp{y}^{\{y\}} ) =$
$\MU y.\; \const{C}( y )$.
The $\mu$-term $t$ of the form $\MU x.\; \const{C}( \MU y.\; \const{C}( x ) )$ is also abnormal,
since $\MU y.\; \const{C}( x )$ is a bisimilar subterm of $t$,
noting that
$t\interp{\MU y.\; \const{C}( x )}^\emptyset =
\MU y.\; \const{C}( t\interp{x}^{\{y\}} ) =
\MU y.\; \const{C}( t\interp{\MU x.\; \const{C}( \MU y.\; \const{C}( x ) ) }^{\{y\}} ) =
\MU y.\; \const{C}( \MU x.\; \const{C}( t\interp{\MU y.\; \const{C}( x )}^{\{x,y\}} ) ) =
\MU y.\allowbreak\; \const{C}( \MU x.\; \const{C}( y ) )$
is $\alpha$-equivalent to $t\interp{x}^\emptyset = t$.
% for the renaming $\{ x \mapsto y, y \mapsto x \}$.

For any $\mu$-term $t$ of the form $\MU x.\; \const C( \vec u )$, we can
recursively construct its \emph{normal form} $\nf{t}$,
by replacing all of the bisimilar subterms of $t$ with $x$
%and by replacing all of the abnormal subterms of $t$ with their corresponding normal forms.
and by normalizing $t$'s subterms.
For example, $\nf{\MU x.\; \const{C}( \MU y.\; \const{C}( x ) )} = \MU x.\; \const{C}( x )$.

\begin{lemma}
\label{lem:mu-norm-arg}
If $\vec u$ are in normal form and $t = \nf{\MU x.\; \const{C}( \vec u )} = \MU x.\; \const{C}( \vec w )$,
then $\vec u \vsim t\interp{\vec w}$.
\end{lemma}
\begin{proof}
\rem{TODO}
\end{proof}

@@

%\begin{lemma}
%\label{lem:mu-cong}
%If $t_u = \MU x.\; \const{C}( \vec u )$ and $t_w = \MU y.\; \const{C}( \vec w )$ are in normal form,
%then $t_u\interp{ \vec u } \vsim t_w\interp{ \vec w }$ if and only if $t_u \vsim t_w$.
%\end{lemma}
%\begin{proof}
%\rem{TODO}
%\end{proof}

We now define a class of models that we consider for a saturated set of equalities~$\Ec$.
Let $\M( \tau )$ denote the interpretation type $\tau$ in $\M$,
that is, a set of domain elements for that type.
For a term $t$, let $\M( t )$ denote the interpretation of $t$ in $\M$,
that is, an element of $\M( \tau )$ where $\tau$ is the type of $t$.
\rem{Clarify that non-codatatype terms are interpreted as $\mu$-terms, although dropping the $\mu$'s gives the standard representation?
(GOOD IDEA. A BIT OF REDUNDANCY CANNOT HURT.)}

\begin{definition}[Normal Model]
\afterDot%
\label{def:norm-model}%
\rm
A model $\M$ is \emph{normal} if these conditions are met:
\begin{enumerate}
\item
For all types $\tau$,
$\M( \tau )$ is a maximal set of closed normal $\mu$-terms of that type that are
unique up to renaming of $\mu$-bound variables,
and acyclic when $\tau \not\in \Codata$.
\item
For all constructor terms $\const{C}( \vec t \,)$ of type $\tau$,
$\M( \const{C}( \vec t \,) )$ is the value
in $\M( \tau )$ that is $\alpha$-equivalent to
$\nf{\MU x.\; \const{C}( \M( \vec t \,) )}$ where $x$ is fresh.
\item
For all selector terms $\const s^k_j( t )$ of type $\tau$, % for some $\vec u$,
$\M( \const s^k_j( t ) )$ is the value
in $\M( \tau )$ that is $\alpha$-equivalent to
$\M( t )\interp{u_j}$
when $\M( t )$ is of the form $\MU x.\; \const{C}_j( \vec u )$.
\end{enumerate}
\end{definition}

@@

We first state the following properties of $\Val^\ast$,
where we write $\muvar( t )$ to denote $t$ if $t$ is a variable, or ${\Varec{x}}$ if $t$ is of the form $\MU {\Varec{x}}.\; u$.

\begin{lemma}
\label{lem:model-completion}
If $\Val$ is constructed for a saturated set $\Ec$
and $\Val^\ast$ is a completion of $\Val$ for normal model $\M$, the following holds:
\begin{enumerate}
\item[\rm 1.] $\Val^\ast \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$, where $\ec{x}$ has type $\tau$.
\item[\rm 2.]
$\Val^\ast \ec{x} = \Val^\ast \ec{y} \interp{ t }$ for
all subterms $t$ of $\Val^\ast \ec{y}$ with $\muvar(t) = {\Varec{x}}$. 
\item[\rm 3.]
$\Val^\ast \ec{x} \vsim \Val^\ast \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{proof}
To show 1, we first show that $\Val^\ast$ contains no free variables.
Assume by contradiction that $\Val^\ast$ contained a free variable ${\Varec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else ${\Varec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then $\tau$ is corecursive by assumption, and by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be $1$.
Since \rn{Singleton} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}^\M_{\Varec{y}}(\Val^\ast)$ of type $\tau$.
This is a contradiction, since our model completion may assign the value in the domain of $\tau$ to ${\Varec{y}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction,
since there are only a finite number of closed terms in $\mathcal{T}^\M_{\Varec{y}}(\Val^\ast)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_{\Varec{y}}(\Val^\ast)$ to ${\Varec{y}}$.
By construction, $\Val^\ast \ec{x}$ is normal.
Since \rn{Acyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\Val^\ast$ applies substitutions of the form
$\{ {\Varec{y}} \mapsto t \}$, where $t$ is acyclic when the type of ${\Varec{y}}$ is not a codatatype.
Thus, $\Val^\ast \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\Val^\ast \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$.

@@

Thus, by induction on the number of applications of the above rule used to obtain $\Val^\ast$,
we have that $\Val^\ast$ satisfies statements 2 and 3.
\qed
\end{proof}

>>>>

Given a $\mu$-term $t$ of the form $\MU x.\; \const{C}( t_1, \ldots, t_n )$ with subterm $u$,
we say $u$ is a \emph{bisimilar subterm} of $t$ if it is of the form
$\MU y.\; \const{C}( u_1, \ldots, u_n )$ and $t\interp{ t_k } \aequiv t\interp{ u_k }$ for $k = 1,\ldots, n$.
The term $t$ is \emph{normal} if it does not contain a strict, bisimilar subterm
and all of its strict subterms are also normal.
For example, the $\mu$-term $t = \MU x.\; \const{C}( \MU y.\; \const{C}( y ) )$ is abnormal,
since $\MU y.\; \const{C}( y )$ is a bisimilar subterm of $t$.
In particular, note that their arguments have the same interpretation in $t$,
that is,
$t\interp{\MU y.\; \const{C}( y )}^\emptyset =$
$\MU y.\; \const{C}( t\interp{y}^{\{y\}} ) =$
$\MU y.\; \const{C}( y )$
is $\alpha$-equivalent to
$t\interp{y}^\emptyset =$
$\MU y.\; \const{C}( t\interp{y}^{\{y\}} ) =$
$\MU y.\; \const{C}( y )$.
The $\mu$-term $t$ of the form $\MU x.\; \const{C}( \MU y.\; \const{C}( x ) )$ is also abnormal,
since $\MU y.\; \const{C}( x )$ is a bisimilar subterm of $t$,
noting that
$t\interp{\MU y.\; \const{C}( x )}^\emptyset =
\MU y.\; \const{C}( t\interp{x}^{\{y\}} ) =
\MU y.\; \const{C}( t\interp{\MU x.\; \const{C}( \MU y.\; \const{C}( x ) ) }^{\{y\}} ) =
\MU y.\; \const{C}( \MU x.\; \const{C}( t\interp{\MU y.\; \const{C}( x )}^{\{x,y\}} ) ) =
\MU y.\allowbreak\; \const{C}( \MU x.\; \const{C}( y ) )$
is $\alpha$-equivalent to $t\interp{x}^\emptyset = t$.
% for the renaming $\{ x \mapsto y, y \mapsto x \}$.
For any $\mu$-term $t$ of the form $\MU x.\; \const C( \vec u )$, we can
recursively construct its \emph{normal form} $\nf{t}$,
by replacing all of the bisimilar subterms of $t$ with $x$
%and by replacing all of the abnormal subterms of $t$ with their corresponding normal forms.
and by normalizing $t$'s subterms.
For example, $\nf{\MU x.\; \const{C}( \MU y.\; \const{C}( x ) )} = \MU x.\; \const{C}( x )$.

\begin{report}
\begin{lemma}
\label{lem:mu-norm-arg}
If $\vec u$ are in normal form and $t = \nf{\MU x.\; \const{C}( \vec u )} = \MU x.\; \const{C}( \vec w )$,
then $\vec u \vsim t\interp{\vec w}$.
\end{lemma}
\begin{proof}
\rem{TODO}
\end{proof}

@@

%\begin{lemma}
%\label{lem:mu-cong}
%If $t_u = \MU x.\; \const{C}( \vec u )$ and $t_w = \MU y.\; \const{C}( \vec w )$ are in normal form,
%then $t_u\interp{ \vec u } \vsim t_w\interp{ \vec w }$ if and only if $t_u \vsim t_w$.
%\end{lemma}
%\begin{proof}
%\rem{TODO}
%\end{proof}
\end{report}

We now define a class of models that we consider for a saturated set of equalities~$\Ec$.
Let $\M( \tau )$ denote the interpretation type $\tau$ in $\M$,
that is, a set of domain elements for that type.
For a term $t$, let $\M( t )$ denote the interpretation of $t$ in $\M$,
that is, an element of $\M( \tau )$ where $\tau$ is the type of $t$.
In the following,
\emph{all} types are interpretated as a set of $\mu$-terms.
This is done to simplify the presentation.
Since all values of types in $\Data$ and $\Nondata$ are acyclic,
a typical representation for these values can be obtained by
dropping each of its $\mu$-bindings. 
For instance, the $\mu$-term $\mu x. \const{0}$ representing the integer constant zero
is simply $\const{0}$ after dropping the $\mu$-binding of $x$.

\begin{definition}[Normal Model]
\afterDot%
\label{def:norm-model}%
\rm
A model $\M$ is \emph{normal} if these conditions are met:
\begin{enumerate}
\item
For all types $\tau$,
$\M( \tau )$ is a maximal set of closed normal $\mu$-terms of that type that are
unique up to renaming of $\mu$-bound variables,
and acyclic when $\tau \not\in \Codata$.
\item
For all constructor terms $\const{C}( \vec t \,)$ of type $\tau$,
$\M( \const{C}( \vec t \,) )$ is the value
in $\M( \tau )$ that is $\alpha$-equivalent to
$\nf{\MU x.\; \const{C}( \M( \vec t \,) )}$ where $x$ is fresh.
\item
For all selector terms $\const s^k_j( t )$ of type $\tau$, % for some $\vec u$,
$\M( \const s^k_j( t ) )$ is the value
in $\M( \tau )$ that is $\alpha$-equivalent to
$\M( t )\interp{u_k}$
when $\M( t )$ is of the form $\MU x.\; \const{C}_j( \vec u )$.
\end{enumerate}
\end{definition}

@@

We first state the following properties of $\Val^\ast$,
where we write $\muvar( t )$ to denote $t$ if $t$ is a variable, or ${\Varec{x}}$ if $t$ is of the form $\MU {\Varec{x}}.\; u$.
\begin{paper} 
The proof of this lemma can be found in the extended version of this paper.
\end{paper}

\begin{lemma}
\label{lem:model-completion}
If $\Val$ is constructed for a saturated set $\Ec$
and $\Val^\ast$ is a completion of $\Val$ for normal model $\M$, the following holds:
\begin{enumerate}
\item[\rm 1.] $\Val^\ast \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$, where $\ec{x}$ has type $\tau$.
\item[\rm 2.]
$\Val^\ast \ec{x} = \Val^\ast \ec{y} \interp{ t }$ for
all subterms $t$ of $\Val^\ast \ec{y}$ with $\muvar(t) = {\Varec{x}}$. 
\item[\rm 3.]
$\Val^\ast \ec{x} \vsim \Val^\ast \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{report}
\begin{proof}
To show 1, we first show that $\Val^\ast$ contains no free variables.
Assume by contradiction that $\Val^\ast$ contained a free variable ${\Varec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else ${\Varec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then $\tau$ is corecursive by assumption, and by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be $1$.
Since \rn{Singleton} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}^\M_{\Varec{y}}(\Val^\ast)$ of type $\tau$.
This is a contradiction, since our model completion may assign the value in the domain of $\tau$ to ${\Varec{y}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction,
since there are only a finite number of closed terms in $\mathcal{T}^\M_{\Varec{y}}(\Val^\ast)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_{\Varec{y}}(\Val^\ast)$ to ${\Varec{y}}$.
By construction, $\Val^\ast \ec{x}$ is normal.
Since \rn{Acyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\Val^\ast$ applies substitutions of the form
$\{ {\Varec{y}} \mapsto t \}$, where $t$ is acyclic when the type of ${\Varec{y}}$ is not a codatatype.
Thus, $\Val^\ast \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\Val^\ast \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$.

@@

Thus, by induction on the number of applications of the above rule used to obtain $\Val^\ast$,
we have that $\Val^\ast$ satisfies statements 2 and 3.
\qed
\end{proof}
\end{report}
